A tool set for random number generation on GPUs in R

### 2021-11-13 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
update paper 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2961 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

}//kernel
\end{lstlisting}
\end{framed}
This kernel executes over a 2-dimensional NDRange. Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers. In each grid of (a) that represents a work-item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work-item. Each stream object is passed to each work-item. The number in a cell of the matrix corresponds to the stream index, it indicates which stream generates a random number for this matrix cell. %For example, stream 0 which is allocated in work-item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
\begin{figure}[H]
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f1} }}
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f2} }}
    \caption{Using streams to produce a $4 \times 6$ matrix of random U(0,1) numbers. \label{fig:1}}
\end{figure}

Now we use the streams created in section \ref{createstreams} to generate two vectors of double-precision i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
<<simStuff2, eval=TRUE,tidy=TRUE>>=
as.vector(clrng::runif(n=6, streams=myStreamsGpu, Nglobal=c(1,nrow(myStreamsGpu)), type = "double"))
@
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: verbose print extra information if >1.
\end{itemize}
\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu} and \code{myStreamsGpu2} here. Reusing \code{myStreamsGpu} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu} after it being used.
<<eval=TRUE,tidy=TRUE, include=FALSE>>=
t(matrix(as.matrix(myStreamsGpu), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{saveRDS} function, and later recall the streams with the \fct{readRDS} function, in this way we can reproduce the exact same sequence of random numbers in simulations. Below is the code that saves streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
<<save streams,echo=TRUE,eval=TRUE>>=
saveRDS(as.matrix(createStreamsGpu(n=4)), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
@

@@

      // cos for local0, sine for local1
      sinOrCosPart1 = cos(part[1] + addForSine);
      temp = part[0]*sinOrCosPart1;
        if(DrowInBounds) out[Dentry] = temp;
      barrier(CLK_LOCAL_MEM_FENCE);

  }//Dcol
}//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}//kernel
\end{lstlisting}
\end{framed}
We illustrate in Figure \ref{fig3} how this kernel executes in practice with the previous toy example: to create a $4 \times 6$ matrix of Gaussian random numbers using a $2 \times 4$ NDRange. %Each stream is copied to the private memory of each work-item. 
Work-items are organized to be in 1 by 2 work-groups as the random numbers are generated in pairs using the Box-Muller method. The shading on the grids in the NDRange distinguishes work-groups. The numbers in the third row in each gird of NDRange represents the local index of each work-item. Each work-group generates a pair of Gaussian random numbers \code{(X,Y)} in each iteration, where work-item of `\code{get_local_id(1)}$=0$' generates the \code{X} of a pair, and work-item of `\code{get_local_id(1)}$=1$' generates the \code{Y} of a pair. %Here with the loop iterating over columns, the cells of matrix are filled from left to right, that is, in the first iteration, the 8 work-items generate random numbers in the leftmost two columns (8 cells) of the matrix simultaneously, the second iteration fills random numbers in the middle two columns of the matrix, and the third iteration fills the rest two columns of the matrix simultaneously. %Therefore, each work-item generates one Gaussian random number in each iteration, 
The matrix filling process is the same as that shown in Figure \ref{fig:1}.

\begin{figure}[H]%
    \centering
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f3} }}%
    \qquad
    \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f4} }}%
    \caption{Using streams to produce a $4 \times 6$ matrix of normal random numbers.}%
    \label{fig3}
\end{figure}

>>>>

}//kernel
\end{lstlisting}
\end{framed}
%Figure \ref{fig:1} illustrates how this kernel executes in practice with a simple example. Suppose we have a $2 \times 4$ NDRange ((a) in Figure \ref{fig:1}) and want to create a $4 \times 6$ matrix ((b) in Figure \ref{fig:1}) of uniform random numbers. In each grid of (a) that represents a work-item, there are two rows of numbers, the number in the above row is the stream index (from 0 to 7), and a pair of numbers in the second row is the 2-dimensional global ID of the work-item. Each stream object is passed to each work-item. The number in a cell of the matrix corresponds to the stream index, it indicates which stream generates a random number for this matrix cell. %For example, stream 0 which is allocated in work-item $(0,0)$ generates a random number in matrix $(0,0)$ in the first iteration, then moves on to matrix $(0,2)$ and generates a random number there in the second iteration, in the third iteration it generates a random number in matrix $(0,4)$ and so on until it goes out of the matrix range in the next iteration.
% \begin{figure}[H]
%     \centering
%     \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f1} }}
%     \qquad
%     \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f2} }}
%     \caption{Using streams to produce a $4 \times 6$ matrix of random U(0,1) numbers. \label{fig:1}}
% \end{figure}

Now we use the streams created in section \ref{createstreams} to generate two vectors of double-precision i.i.d. $U(0,1)$ random numbers with the function \fct{runif}. To view the generated random numbers, we need to convert them to R vectors or matrices, by doing this, the random numbers are moved from the global memory of the GPU device to the host.
<<simStuff2, eval=TRUE,tidy=TRUE>>=
as.vector(clrng::runif(n=6, streams=myStreamsGpu, Nglobal=c(1,nrow(myStreamsGpu)), type = "double"))
@
The arguments of the \fct{clrng::runif} are described as follows:
\begin{itemize}
\itemsep0em 
  \item \code{n}: a vector of length 2 specifying the row and column number if to create a matrix, or a number specifying the length if to create a vector.
  \item \code{streams}: streams for random number generation.
  \item \code{Nglobal}: global index space. Default is set as $(64,8)$.
  \item \code{type}: ``double'' or ``float'' or ``integer'' format of generated random numbers.
  \item \code{verbose}: verbose print extra information if >1.
\end{itemize}
%\fct{runif} returns the same sequence of random numbers because it uses the same streams \code{myStreamsGpu} and \code{myStreamsGpu2} here. 
Reusing \code{myStreamsGpu} will produce a vector different from \code{sim_1}. Because each time a random number is generated, the current state of the stream advances by one position. %Below shows the \code{myStreamsGpu} after it being used.
<<eval=TRUE,tidy=TRUE, include=FALSE>>=
t(matrix(as.matrix(myStreamsGpu), nrow(myStreamsCpu), ncol(myStreamsCpu), dimnames = dimnames(myStreamsCpu)))
@
Unlike the objects created by \proglang{R}, by default they remain in the memory after a restart. Streams created on GPU does not remain in the memory when a new \proglang{R} session begins. However, we can save the streams on the CPU in a file using the \fct{saveRDS} function, and later recall the streams with the \fct{readRDS} function, in this way we can reproduce the exact same sequence of random numbers in simulations. Below is the code that saves streams to a data file called \code{streams_from_GPU.rds} on CPU and then load it back and transfer it to GPU.
<<save streams,echo=TRUE,eval=TRUE>>=
saveRDS(as.matrix(createStreamsGpu(n=4)), "myStreams.rds")
# Load the streams object as streams_saved
streams_saved <- vclMatrix(readRDS("myStreams.rds"))
@

@@

      // sine for local[1], cos for local[0]
      sinOrCosPart1 = cos(part[1] + addForSine);
      temp = part[0]*sinOrCosPart1;
      if(DrowInBounds) out[Dentry] = temp;
      barrier(CLK_LOCAL_MEM_FENCE);

  }//Dcol
}//Drow
streamsFromPrivate(streams,g1,g2,startvalue);
}//kernel
\end{lstlisting}
\end{framed}
%We illustrate in Figure \ref{fig3} how this kernel executes in practice with the previous toy example: to create a $4 \times 6$ matrix of Gaussian random numbers using a $2 \times 4$ NDRange. %Each stream is copied to the private memory of each work-item. 
%Work-items are organized to be in 1 by 2 work-groups as the random numbers are generated in pairs using the Box-Muller method. The shading on the grids in the NDRange distinguishes work-groups. The numbers in the third row in each gird of NDRange represents the local index of each work-item. Each work-group generates a pair of Gaussian random numbers \code{(X,Y)} in each iteration, where work-item of `\code{get_local_id(1)}$=0$' generates the \code{X} of a pair, and work-item of `\code{get_local_id(1)}$=1$' generates the \code{Y} of a pair. %Here with the loop iterating over columns, the cells of matrix are filled from left to right, that is, in the first iteration, the 8 work-items generate random numbers in the leftmost two columns (8 cells) of the matrix simultaneously, the second iteration fills random numbers in the middle two columns of the matrix, and the third iteration fills the rest two columns of the matrix simultaneously. %Therefore, each work-item generates one Gaussian random number in each iteration, The matrix filling process is the same as that shown in Figure \ref{fig:1}.

% \begin{figure}[H]%
%     \centering
%     \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f3} }}%
%     \qquad
%     \subfloat[]{{\includegraphics[width=8cm]{/home/ruoyong/diseasemapping/pkg/gpuRandom/inst/documents/paper1_2021_5/f4} }}%
%     \caption{Using streams to produce a $4 \times 6$ matrix of normal random numbers.}%
%     \label{fig3}
% \end{figure}
